---
title: "SbCompendium: Example Workflow"
author: "Priscilla Doupnik"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{SbCompendium: Example Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(SbCompendium)
library(reticulate)
```


# Introduction

SbCompendium provides tools for analyzing sorghum RNA-seq datasets from the Mullet lab.
This vignette walks through a complete example workflow using a small example dataset
included with the package.

This workflow covers:

- Importing transcript-level data from the Mullet Sorghum RNA-seq datasets and aggregating transcripts to genes
- UMAP quality control of sample relationships
- Differential expression analysis using DESeq2
- k-means clustering of expression patterns
- Exporting heatmap tables with optional annotations and clustering results

## Example dataset

The package includes a small simulated RNA-seq dataset containing:

- transcript IDs
- replicate-level TPM expression
- replicate-level count data
- condition-level TPMmean expression

This dataset is designed to work with all functions in the package.
More details and options for each function can be found in the function's manual.

```{r}
data("example_data")
```

In this example, the compendium contains two tables: the TX08001 experimental table (nodal_buds_28) with reps from nodal root buds from Phytomers 7, 10, 15, and 21 and the annotation table (sorghum_gene_annotation table).

```{r}
list_compendium(example_data)
```


# Import transcript-level data as gene-level expression

RNA-seq quantification files often contain transcript-level expression.
The `importTable` function aggregates transcripts to gene-level values
by summing expression across transcripts belonging to the same gene.

```{r}
gene_expr <- importTable(example_data[["nodal_buds_28"]])
head(gene_expr)
```


# UMAP visualization

`run_umap_tpm` can be used to visualize sample relationships using replicate-level TPM
expression and determine if there are any outliers.
In this example, each of the nodal root bud groups cluster near one another. 
We can also easily seed that group 4, node 21 only has 2 reps instead of 3
In addition, the younger nodes (15 and 21) cluster closer to each other than the older nodes
```{r}
um <- run_umap_tpm(gene_expr, n_neighbors = 3)

#head(um$umap_df)

plot_umap(um)
```

## Reorder Group labels if necessary

Group ordering can be adjusted for plotting using `set_umap_group_order`.

In this case, they are already correct, so let's put them in the wrong order and then fix it.
```{r}
#First print the group names:
print_umap_groups(um)

#then reorder them - #this is now the wrong order
um2 <- set_umap_group_order(
  um,
c(
  "TX08001_Nodal_Root_Bud_Node_10",
  "TX08001_Nodal_Root_Bud_Node_15",
  "TX08001_Nodal_Root_Bud_Node_7",
  "TX08001_Nodal_Root_Bud_Node_21"
)
)

plot_umap(um2)
```
lets reorder them properly now
```{r}
um2 <- set_umap_group_order(
  um,
c(
  "TX08001_Nodal_Root_Bud_Node_7",
  "TX08001_Nodal_Root_Bud_Node_10",
  "TX08001_Nodal_Root_Bud_Node_15",
  "TX08001_Nodal_Root_Bud_Node_21"
)
)

plot_umap(um2)
```


# Build a DESeq2 object

The `build_deseq` function constructs a DESeq2 dataset from the
expression table.

Genes can optionally be filtered using a TPM threshold.

`dds_out` will include:
- the DESeq2 object (dds)
- a subset of expression values (subset_df)
- column name helpers (tpm_cols, counts_cols)
- grouping information (groups, sample_names, rep_counts)
- sample_table used in DEseq analysis
- colData used in DEseq analysis
- chosen TPM threshold (TPM_thresh)

```{r}
i <- 2 

data <- sorghum_compendium[[i]]

gene_expr <- importTable(sorghum_compendium[[i]])

dds_out <- build_deseq(
  gene_expr,
  TPM_thresh = 5,
  run_deseq = TRUE
)

#inspect sample information
dds_out$sample_table
dds_out$groups
#number of replicates per condition
dds_out$rep_counts


colnames(data) <- gsub(
  "(Core|Rind)_TPM$",
  "\\1_TPMmean",
  colnames(data)
)

sorghum_compendium[[i]] <- data
```

# Run a differential expression contrast

## A single DESeq2 contrast
The `run_contrast()` function runs a single DESeq2 contrast between two groups of the user's choice.

If one does not wish to cluster the results into similar expression groups, then use the results, contrast$top_de, in the `build_heatmap` function and export that dataframe using the `write_heatmap_xlsx` function

```{r}
#find the group names

contrast <- run_contrast(
  dds_out$dds,
  groupA = "TX08001_Nodal_Root_Bud_Node_7",
  groupB = "TX08001_Nodal_Root_Bud_Node_21"
)

#Top differentially expressed genes:
head(contrast$top_de)
```

## Run all pairwise contrasts

The `run_all_pairwise_contrasts()` function computes DESeq2 results for all
pairwise comparisons among condition levels.
This step can take a while depending on how many total constrasts it must compute

If one does not wish to cluster the results into similar expression groups, then use the results, pairwise$top_by_gene, in the `build_heatmap` function and export that data frame using the `write_heatmap_xlsx` function

```{r}
pairwise <- run_all_pairwise_contrasts(
  dds_out$dds,
  factor_name = "condition",
  run_deseq = FALSE,   # already ran DESeq in build_deseq()
  alpha = 0.05,
  fc_thresh = 3,       # fold-change threshold (FC scale)
  add_fc = TRUE,
  filter_significant = TRUE
)

# show which contrasts were tested
pairwise$contrasts

# combined significant results (long format)
head(pairwise$all_results)

# per-gene "best" contrast summary
head(pairwise$top_by_gene)
```

# Kmeans Clustering
## Prepare expression data for clustering

K-means clustering is typically performed using TPMmean expression values
for the genes of interest.

The `prepare_kmeans_matrix_topDE_TPMmean` function builds the expression
matrix used for clustering.

Either `pairwise$top_by_gene` or `contrast$top_de` can be used for clustering

we'll use `pairwise$top_by_gene` for our example
```{r}
mat_pairwise <- prepare_kmeans_matrix_topDE_TPMmean(
  top_de = pairwise$top_by_gene,
  gene_expr = gene_expr
)

dim(mat_pairwise)
```

## Determine the optimal number of clusters

The `evaluate_kmeans_k` function helps determine the best number of clusters
using silhouette and WSS diagnostics.

4 clusters is the optimal number in both graphs
```{r}
evaluate_kmeans_k(mat_pairwise, k_max = 10)
```

## Run k-means clustering

Once a cluster number is chosen, run k-means using the `run_kmeans` function and determine which genes are in which cluster:

```{r}
km <- run_kmeans(mat_pairwise, centers = 4)

head(km$clusters)
```
# Export results
## Build a heatmap table to export

The `build_heatmap` function combines:

- TPMmean expression
- cluster assignments
- differential expression statistics
- gene annotations

into a single table suitable for heatmap visualization.
We will look at all nodal root buds but if we wanted to view only some, we would use:  groups = c("TX08001_Nodal_Root_Bud_Node_7", "TX08001_Nodal_Root_Bud_Node_21")

```{r}
heatmap_df <- build_heatmap(
  gene_expr = gene_expr,
  clusters = km,
  top_de = pairwise$top_by_gene,
  fc_cols = c( "log2FoldChange","contrast", "padj", "log2FoldChange"),
  annot_df = example_data$Sorghum_gene_annotation
)

head(heatmap_df)
```

## Export heatmap dataframe

The `write_heatmap_xlsx` function writes the heatmap table to Excel
using built-in color rules for the TPMmean values.

The function is given below but commented out. 
The output file can be found in the extdata folder and details are shown below

The heatmap Excel file contains colors and frozen panes. 
To see these properly, open the file in Excel or LibreOffice."
```{r}
# write_heatmap_xlsx(
#   heatmap_df,
#   "heatmap_export.xlsx"
# )

# Heatmap Excel demo
heatmap_demo_file <- system.file(
  "extdata/heatmap_export_demo.xlsx",
  package = "SbCompendium"
)

if (file.exists(heatmap_demo_file)) {
  print("Heatmap Excel demo file (conditional formatting + frozen panes):")
  print(heatmap_demo_file)
} else {
  warning("Heatmap demo file not found! Make sure it's in inst/extdata/")
}
```

# Export cluster gene lists for DAVID GO term analysis

Cluster gene lists can be exported in a DAVID-compatible format. 
The `export_clusters_david` function is given below commented out. 
The output files can be accessed below
```{r}

# export_clusters_david(
#   heatmap_df,
#   out_dir = "inst/examples/cluster_gene_lists"
# )

# Cluster gene list demos
cluster_demo_dir <- system.file(
  "extdata/cluster_gene_lists_demo", 
  package = "SbCompendium"
)

cluster_files <- list.files(cluster_demo_dir, full.names = TRUE)

if (length(cluster_files) > 0) {
  print("Cluster gene list demo files (ready for DAVID GO term analysis):")
} else {
  warning("No cluster demo files found! Make sure they're in inst/extdata/cluster_gene_lists_demo/")
}

# -----------------------------
# Preview a small portion in R
# -----------------------------
# Show first few rows of cluster 1 (if available)
if (length(cluster_files) >= 1) {
  cluster1 <- read.delim(cluster_files[1], header = FALSE)
  head(cluster1)
}
```

