---
title: "SbCompendium: Example Workflow"
author: "Priscilla Doupnik"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{SbCompendium: Example Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(SbCompendium)
library(reticulate)
```


# Introduction

SbCompendium provides tools for analyzing sorghum RNA-seq datasets from the Mullet lab.
This vignette walks through a complete example workflow using a small example dataset
included with the package.

This workflow covers:

- Importing transcript-level data from the Mullet Sorghum RNA-seq datasets and aggregating transcripts to genes
- UMAP quality control of sample relationships
- Differential expression analysis using DESeq2
- k-means clustering of expression patterns
- Exporting heatmap tables with optional annotations and clustering results

## Example dataset

The package includes a small simulated RNA-seq dataset containing:

- transcript IDs
- replicate-level TPM expression
- replicate-level count data
- condition-level TPMmean expression

This dataset is designed to work with all functions in the package.
More details and options for each function can be found in the function's manual.

```{r}
data("example_data")
```

In this example, the compendium contains two tables: the TX08001 experimental table (nodal_buds_28) with reps from nodal root buds from Phytomers 7, 10, 15, and 21 and the annotation table (sorghum_gene_annotation table).

```{r}
list_compendium(example_data)
```


# Import transcript-level data as gene-level expression

RNA-seq quantification files often contain transcript-level expression.
The `importTable` function aggregates transcripts to gene-level values
by summing expression across transcripts belonging to the same gene.

```{r}
gene_expr <- importTable(example_data[["nodal_buds_28"]])
head(gene_expr)
```


# UMAP visualization

`run_umap_tpm` can be used to visualize sample relationships using replicate-level TPM
expression and determine if there are any outliers.
In this example, each of the nodal root bud groups cluster near one another. 
We can also easily see that group 4, node 21, has two replicates instead of three.
In addition, the younger nodes (15 and 21) cluster closer to each other than the older nodes.

Setting `seed` makes the layout reproducible. Here, genes must exceed a TPM of
5 in at least one sample, and `return_input = TRUE` retains the filtered,
gene-by-sample input matrix in the returned object.
```{r}
um <- run_umap_tpm(
  gene_expr,
  TPM_thresh = 5,
  n_neighbors = 3,
  seed = 20,
  return_input = TRUE
)

# Inspect coordinates, inferred sample groups, replicate counts, and settings
head(um$umap_df)
um$sample_table
um$rep_counts
um$meta
dim(um$input_matrix)

```



### Default labels

Calling `plot_umap()` without color arguments displays black numeric labels and
a legend mapping those labels to the full group names.

```{r}
plot_umap(um2, label_type = "group", text_size = 5)
```

### Sample labels colored by group

With `label_type = "sample"`, each observation is labeled with its sample-column
name without the `_TPM` suffix. Labels are colored by their inferred group.

```{r}
plot_umap(
  um2,
  label_type = "sample",
  text_size = 5
)
```

### Points colored by group

With `label_type = "color"`, samples are drawn as colored points rather than
text labels.

```{r}
plot_umap(
  um2,
  label_type = "color"
)
```


## Reorder Group labels if necessary

Group ordering can be adjusted for plotting using `set_umap_group_order`.

In this case, they are already correct, so let's put them in the wrong order and then fix it.
```{r}
#First print the group names:
print_umap_groups(um)

#then reorder them - #this is now the wrong order
um2 <- set_umap_group_order(
  um,
c(
  "TX08001_Nodal_Root_Bud_Node_10",
  "TX08001_Nodal_Root_Bud_Node_15",
  "TX08001_Nodal_Root_Bud_Node_7",
  "TX08001_Nodal_Root_Bud_Node_21"
)
)

plot_umap(um2)
```
Let's reorder them properly and then compare the plotting options. The numeric
labels are assigned from the reordered factor levels, so label 1 corresponds to
Node 7, label 2 to Node 10, and so on.
```{r}
um2 <- set_umap_group_order(
  um,
c(
  "TX08001_Nodal_Root_Bud_Node_7",
  "TX08001_Nodal_Root_Bud_Node_10",
  "TX08001_Nodal_Root_Bud_Node_15",
  "TX08001_Nodal_Root_Bud_Node_21"
)
)
```

# Build a DESeq2 object

The `build_deseq` function constructs a DESeq2 dataset from the
expression table.

Genes can optionally be filtered using a TPM threshold.

`dds_out` will include:
- the DESeq2 object (dds)
- a subset of expression values (subset_df)
- column name helpers (`tpm_cols`, `count_cols`)
- grouping information (groups, sample_names, rep_counts)
- `sample_table` used in DESeq2 analysis
- `colData` used in DESeq2 analysis
- chosen TPM threshold (TPM_thresh)

```{r}
dds_out <- build_deseq(
  gene_expr,
  TPM_thresh = 5,
  run_deseq = TRUE
)

#inspect sample information
dds_out$sample_table
dds_out$groups
#number of replicates per condition
dds_out$rep_counts

```

# Inspect samples with PCA and heatmaps

The `plot_deseq_qc()` function performs a variance-stabilizing transformation
and produces four quality-control views after the initial DESeq2 run:

- a PCA with sample labels and percent variance on each axis
- the same PCA with points only for a less crowded view
- a sample-to-sample Pearson correlation heatmap
- a row-scaled expression heatmap annotated by condition

The function accepts the complete result from `build_deseq()` and returns the
transformed data, plots, PCA data, and correlation matrix for further use.

```{r deseq-qc-plots, fig.width=14, fig.height=10, out.width="100%"}
qc <- plot_deseq_qc(
  dds_out$dds,
  intgroup = "condition",
  blind = TRUE,
  cor_method = "pearson",
  heatmap_scale = "row",
  show_rownames = FALSE,
  cor_cellheight = 20
)

# Inspect or reuse returned results
head(qc$pca_data)
qc$percent_var
qc$cor_matrix

# Reprint either PCA version as needed
qc$pca_plot
qc$pca_plot_points
grid::grid.newpage()
print(qc$correlation_heatmap)
grid::grid.newpage()
print(qc$expression_heatmap)
```

# Run a differential expression contrast

## A single DESeq2 contrast
The `run_contrast()` function runs a single DESeq2 contrast between two groups of the user's choice.

The `lfc_thresh` argument is on the **log2 fold-change scale**. For example,
`lfc_thresh = 2` requires an absolute log2 fold change greater than 2.

If one does not wish to cluster the results into similar expression groups, then use the results, contrast$top_de, in the `build_heatmap` function and export that dataframe using the `write_heatmap_xlsx` function

```{r}
#find the group names

contrast <- run_contrast(
  dds_out$dds,
  groupA = "TX08001_Nodal_Root_Bud_Node_7",
  groupB = "TX08001_Nodal_Root_Bud_Node_21",
  alpha = 0.05,
  lfc_thresh = 2.32
)

#Top differentially expressed genes:
head(contrast$top_de)
```

## Run all pairwise contrasts

The `run_all_pairwise_contrasts()` function computes DESeq2 results for all
pairwise comparisons among condition levels.
This step can take a while depending on how many total contrasts it must compute.

Unlike `run_contrast()`, `fc_thresh` is on the ordinary fold-change scale.
For example, `fc_thresh = 3` retains changes greater than threefold in either
direction by applying `abs(log2FoldChange) > log2(3)` internally.

If one does not wish to cluster the results into similar expression groups, then use the results, pairwise$top_by_gene, in the `build_heatmap` function and export that data frame using the `write_heatmap_xlsx` function

```{r}
pairwise <- run_all_pairwise_contrasts(
  dds_out$dds,
  factor_name = "condition",
  run_deseq = FALSE,   # already ran DESeq in build_deseq()
  alpha = 0.05,
  fc_thresh = 3,       # fold-change threshold (FC scale)
  add_fc = TRUE,
  filter_significant = TRUE
)

# show which contrasts were tested
pairwise$contrasts

# combined significant results (long format)
head(pairwise$all_results)

# per-gene "best" contrast summary
head(pairwise$top_by_gene)
```

# Kmeans Clustering
## Prepare expression data for clustering

K-means clustering is typically performed using TPMmean expression values
for the genes of interest.

The `prepare_kmeans_matrix_topDE_TPMmean` function builds the expression
matrix used for clustering.

Either `pairwise$top_by_gene` or `contrast$top_de` can be used for clustering

we'll use `pairwise$top_by_gene` for our example
```{r}
mat_pairwise <- prepare_kmeans_matrix_topDE_TPMmean(
  top_de = pairwise$top_by_gene,
  gene_expr = gene_expr
)

dim(mat_pairwise)
```

To compare expression patterns for exactly two groups, pass their exact
TPMmean column names through `tpmmean_cols`. These names contain `Node_Phy`,
which differs from the group names inferred from the replicate columns. With
two TPMmean columns, the function avoids row scaling and applies `log2(x + 1)`
by default. This prevents two-column profiles from collapsing to the same
scaled values.

```{r two-group-kmeans, eval=FALSE}
mat_two_groups <- prepare_kmeans_matrix_topDE_TPMmean(
  top_de = contrast$top_de,
  gene_expr = gene_expr,
  tpmmean_cols = c(
    "TX08001_Nodal_Root_Bud_Node_Phy_7_TPMmean",
    "TX08001_Nodal_Root_Bud_Node_Phy_21_TPMmean"
  )
)
```

## Determine the optimal number of clusters

The `evaluate_kmeans_k` function helps determine the best number of clusters
using silhouette and WSS diagnostics.

Four clusters is the optimal number in both graphs. The function returns its
plots and metadata invisibly, so they can also be saved for later inspection.
```{r}
k_eval <- evaluate_kmeans_k(mat_pairwise, k_max = 10, seed = 1)
k_eval$meta
```

## Run k-means clustering

Once a cluster number is chosen, run k-means using the `run_kmeans` function and determine which genes are in which cluster.

Then plot the median expression values across collection points for each cluster.
These can be either raw TPMmean values or log-transformed values.
The trailing `_TPMmean` suffix is removed automatically from the x-axis labels.

```{r}
km <- run_kmeans(mat_pairwise, centers = 4, seed = 1)

head(km$clusters)


p <- plot_cluster_median_profiles_tpm(
    gene_expr = gene_expr,
    clusters = km$clusters,
    log_transform = FALSE,
    title = "Stem development expression clusters"
)

print(p)
```
# Export results
## Build a heatmap table to export

The `build_heatmap` function combines:

- TPMmean expression
- cluster assignments
- differential expression statistics
- gene annotations

into a single table suitable for heatmap visualization.
The first example includes all TPMmean columns.

```{r}
heatmap_df <- build_heatmap(
  gene_expr = gene_expr,
  clusters = km,
  top_de = pairwise$top_by_gene,
  fc_cols = c("FC", "log2FoldChange", "contrast", "padj"),
  annot_df = example_data$Sorghum_gene_annotation
)

head(heatmap_df)
```

The `tpmmean_cols` argument can restrict the exported expression columns to two
conditions while retaining the selected differential-expression and annotation
columns.

```{r}
heatmap_two_groups <- build_heatmap(
  gene_expr = gene_expr,
  clusters = km,
  top_de = pairwise$top_by_gene,
  annot_df = example_data$Sorghum_gene_annotation,
  tpmmean_cols = c(
    "TX08001_Nodal_Root_Bud_Node_Phy_7_TPMmean",
    "TX08001_Nodal_Root_Bud_Node_Phy_21_TPMmean"
  )
)

head(heatmap_two_groups)
```

## Export heatmap dataframe

The `write_heatmap_xlsx` function writes the heatmap table to Excel
using built-in color rules for the TPMmean values.

The write call is commented out because it requires the Python modules listed
in the function documentation. `tempfile()` demonstrates a safe output path
without writing into the package source directory.

The heatmap Excel file contains colors and frozen panes. 
To see these properly, open the file in Excel or LibreOffice.
```{r}
xlsx_file <- tempfile(fileext = ".xlsx")

# Requires pandas, xlsxwriter, and openpyxl in the configured Python environment:
# write_heatmap_xlsx(heatmap_df, xlsx_file)

# Heatmap Excel demo
heatmap_demo_file <- system.file(
  "extdata/heatmap_export_demo.xlsx",
  package = "SbCompendium"
)

if (file.exists(heatmap_demo_file)) {
  print("Heatmap Excel demo file (conditional formatting + frozen panes):")
  print(heatmap_demo_file)
} else {
  warning("Heatmap demo file not found! Make sure it's in inst/extdata/")
}
```

# Export cluster gene lists for DAVID GO term analysis

Cluster gene lists can be exported in a DAVID-compatible format. 
The following example safely writes selected clusters to the R session's
temporary directory.
```{r}
david_dir <- file.path(tempdir(), "david_clusters")

david_files <- export_clusters_david(
  heatmap_df,
  clusters = c(1, 2),
  out_dir = david_dir,
  quiet = TRUE
)

basename(david_files)

# Cluster gene list demos
cluster_demo_dir <- system.file(
  "extdata/cluster_gene_lists_demo", 
  package = "SbCompendium"
)

cluster_files <- list.files(cluster_demo_dir, full.names = TRUE)

if (length(cluster_files) > 0) {
  print("Cluster gene list demo files (ready for DAVID GO term analysis):")
} else {
  warning("No cluster demo files found! Make sure they're in inst/extdata/cluster_gene_lists_demo/")
}

# -----------------------------
# Preview a small portion in R
# -----------------------------
# Show first few rows of cluster 1 (if available)
if (length(cluster_files) >= 1) {
  cluster1 <- read.delim(cluster_files[1], header = FALSE)
  head(cluster1)
}
```

