---
title: "SbCompendium: SummarizedExperiment Workflow"
author: "Priscilla Doupnik"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{SbCompendium: SummarizedExperiment Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(SbCompendium)

resource_candidates <- c(
  file.path("resources", "nodal_buds_28.rds"),
  file.path("..", "resources", "nodal_buds_28.rds")
)
resource_file <- resource_candidates[file.exists(resource_candidates)][1]
have_local_resource <- length(resource_file) == 1 && !is.na(resource_file)

# The development resource is intentionally absent from the package tarball.
# Locally, where it exists, all executable chunks run normally.
knitr::opts_chunk$set(eval = have_local_resource)
```

# Introduction

SbCompendium stores each experiment as a `SummarizedExperiment`. Each object
keeps the data for one experiment together:

- replicate-level TPM values in the `TPM` assay;
- replicate-level counts in the `counts` assay;
- gene annotations in `rowData`;
- sample and experimental metadata in `colData` and `metadata`.

Condition-level mean TPM values are calculated when needed. They are no longer
stored as separate `_TPMmean` columns.

# Load an experiment resource

During local resource development, load the RDS file directly. The
`resources/` directory is deliberately excluded from the package and Git
because ExperimentHub will ultimately distribute these files.

```{r load-resource}
se <- readRDS(resource_file)
se
```

If these chunks are skipped, first run
`Rscript inst/scripts/build_experiment_resources.R`, then knit the vignette
again from the package root. Once the data are published, this loading section
will be replaced by the package's ExperimentHub accessor.

Inspect the assays and metadata before analysis.

```{r inspect-resource}
SummarizedExperiment::assayNames(se)
dim(se)
head(as.data.frame(SummarizedExperiment::colData(se)))
head(as.data.frame(SummarizedExperiment::rowData(se)))
S4Vectors::metadata(se)
```

# Calculate condition means

`summarize_tpm()` replaces the old stored `_TPMmean` columns. By default it
averages the `TPM` assay using the `condition` field in `colData`.

```{r summarize-tpm}
mean_tpm <- summarize_tpm(se)
dim(mean_tpm)
colnames(mean_tpm)
mean_tpm[1:5, 1:4]
```

If another `colData` field defines the desired groups, pass its name through
`group_by`.

```{r alternate-group, eval=FALSE}
mean_by_tissue <- summarize_tpm(se, group_by = "condition")
```

# UMAP visualization

`run_umap_tpm()` now reads replicate-level expression from the `TPM` assay and
sample labels from `colData`. It does not infer biological groups from column
name suffixes when given a `SummarizedExperiment`.

```{r run-umap, fig.width=8, fig.height=6}
um <- run_umap_tpm(
  se,
  group_by = "condition",
  TPM_thresh = 5,
  n_neighbors = 3,
  seed = 20,
  return_input = TRUE
)

head(um$umap_df)
um$rep_counts
plot_umap(um, label_type = "color")
```

The plotting order can still be changed without rerunning UMAP.

```{r reorder-umap}
condition_order <- unique(as.character(
  SummarizedExperiment::colData(se)$condition
))
um_ordered <- set_umap_group_order(um, condition_order)
plot_umap(um_ordered, label_type = "group", text_size = 4)
```

# Build a DESeq2 object

`build_deseq()` reads the `counts` assay, filters genes using the `TPM` assay,
and passes all sample metadata from `colData` to DESeq2. Setting
`run_deseq = FALSE` is useful for a fast structural test.

```{r build-dds-smoke-test}
dds_unfitted <- build_deseq(
  se,
  TPM_thresh = 5,
  design_formula = ~ condition,
  run_deseq = FALSE
)

dds_unfitted$dds
head(dds_unfitted$sample_table)
dds_unfitted$groups
dds_unfitted$rep_counts
```

Run the full model once before contrasts or DESeq2 quality-control plots.

```{r fit-deseq}
dds_out <- build_deseq(
  se,
  TPM_thresh = 5,
  design_formula = ~ condition,
  run_deseq = TRUE
)
```

# Inspect samples with PCA and heatmaps

```{r deseq-qc-plots, fig.width=12, fig.height=9}
qc <- plot_deseq_qc(
  dds_out$dds,
  intgroup = "condition",
  blind = TRUE,
  cor_method = "pearson",
  heatmap_scale = "row",
  show_rownames = FALSE,
  print_plots = FALSE
)

qc$pca_plot_points
grid::grid.newpage()
print(qc$correlation_heatmap)
```

# Differential expression

Use the exact condition values shown by `dds_out$groups`. This example compares
phytomers 7 and 21.

```{r single-contrast}
group_p7 <- "28_TX08001_devo_v_Root_NRB_P7_F"
group_p21 <- "28_TX08001_devo_v_Root_NRB_P21_F"

contrast <- run_contrast(
  dds_out$dds,
  groupA = group_p7,
  groupB = group_p21,
  alpha = 0.05,
  lfc_thresh = 2.32
)

head(contrast$top_de)
```

All pairwise comparisons can be run from the same fitted object. This is more
computationally expensive than a single contrast.

```{r pairwise-contrasts, eval=FALSE}
pairwise <- run_all_pairwise_contrasts(
  dds_out$dds,
  factor_name = "condition",
  run_deseq = FALSE,
  alpha = 0.05,
  fc_thresh = 3,
  add_fc = TRUE,
  filter_significant = TRUE
)

head(pairwise$top_by_gene)
```

# K-means clustering

The clustering preparation function accepts the experiment object directly.
It calculates condition means internally, so `_TPMmean` columns are not needed.

```{r prepare-kmeans}
mat_contrast <- prepare_kmeans_matrix_topDE_TPMmean(
  top_de = contrast$top_de,
  gene_expr = se,
  group_by = "condition"
)

dim(mat_contrast)
```

To cluster only two conditions, pass their exact condition names through
`groups`. With two columns, the function uses `log2(x + 1)` by default instead
of row scaling.

```{r two-condition-kmeans}
mat_two_conditions <- prepare_kmeans_matrix_topDE_TPMmean(
  top_de = contrast$top_de,
  gene_expr = se,
  groups = c(group_p7, group_p21)
)

colnames(mat_two_conditions)
```

Evaluate and fit a clustering solution. `k_max` must be smaller than the number
of genes in the matrix.

```{r run-kmeans, fig.width=8, fig.height=5}
k_max <- min(10, nrow(mat_contrast) - 1L)
if (k_max >= 2) {
  k_eval <- evaluate_kmeans_k(mat_contrast, k_max = k_max, seed = 1)
  k_eval$meta
}

#pick the number of clusters you would like
clusterN <- 2

centers <- min(clusterN, nrow(mat_contrast))
km <- run_kmeans(mat_contrast, centers = centers, seed = 1)
head(km$clusters)
```

Plot cluster profiles using true condition-level mean TPM values derived from
the experiment. Passing the same condition names used by `run_contrast()` to
`tpm_cols` limits the plot to those two conditions instead of displaying every
condition in the experiment.

```{r cluster-profiles, fig.width=10, fig.height=6}
plot_cluster_median_profiles_tpm(
  gene_expr = se,
  clusters = km$clusters,
  tpm_cols = c(group_p7, group_p21),
  group_by = "condition",
  log_transform = FALSE,
  title = "P7 versus P21 expression clusters"
)
```

# Build and export a heatmap table

`build_heatmap()` calculates condition means and, unless `annot_df` is
provided, automatically obtains gene annotations from `rowData(se)`.

```{r build-heatmap}
heatmap_df <- build_heatmap(
  gene_expr = se,
  clusters = km,
  top_de = contrast$top_de,
  fc_cols = c("log2FoldChange", "padj"),
  group_by = "condition"
)

head(heatmap_df)
attr(heatmap_df, "expression_columns")
```

Use `groups` to include only selected condition-mean expression columns.

```{r two-condition-heatmap}
heatmap_two_conditions <- build_heatmap(
  gene_expr = se,
  clusters = km,
  top_de = contrast$top_de,
  fc_cols = c("log2FoldChange", "padj"),
  groups = c(group_p7, group_p21)
)

attr(heatmap_two_conditions, "expression_columns")
```

When the table comes directly from `build_heatmap()`, `write_heatmap_xlsx()`
uses its `expression_columns` attribute. You no longer need to identify
`_TPMmean` columns manually.

```{r export-xlsx, eval=FALSE}
xlsx_file <- file.path(tempdir(), "nodal_buds_heatmap.xlsx")
write_heatmap_xlsx(heatmap_two_conditions, xlsx_file)
xlsx_file
```

# Quick compatibility checks

These assertions provide a short test of the new storage model after editing
the package functions.

```{r compatibility-checks}
stopifnot(
  inherits(se, "SummarizedExperiment"),
  all(c("TPM", "counts") %in% SummarizedExperiment::assayNames(se)),
  nrow(mean_tpm) == nrow(se),
  ncol(mean_tpm) == length(unique(
    SummarizedExperiment::colData(se)$condition
  )),
  identical(colnames(mat_two_conditions), c(group_p7, group_p21)),
  identical(
    attr(heatmap_two_conditions, "expression_columns"),
    c(group_p7, group_p21)
  )
)

message("SummarizedExperiment workflow checks passed.")
```

# Session information

```{r session-info}
sessionInfo()
```
